Bad forecast, good decision: what makes a model relevant for determining policy

### 2020-11-05 manuscript/appendix.Rmd
here we go

---
title: 'Appendix for: "Bad Forecast, Good Decision..."'
date: "`r Sys.Date()`"
author: Carl Boettiger
journal: Appendix
layout: 3p
bibliography: refs.bib
#output:
#  tufte::tufte_handout:
#    dev: cairo_pdf
#    latex_engine: xelatex
output: 
  hrbrthemes::ipsum_pdf:
    dev: cairo_pdf
    latex_engine: xelatex

@@

Here we provide annotated code necessary to completely reproduce all of the analysis
presented in the main paper.
This analysis is run in R [@R] uses `MDPtoolbox` [@MDPtoolbox] for solving 
Markov Decision Processes (MDP) using stochastic dynamic programming functionality,
`expm` for matrix exponentials [@expm], and a few custom MDP functions provided
by our package, `mdplearning` [@mdplearning].  We will also use `tidyverse` 
packages for basic manipulation and plotting [@tidyverse]. This file is also
available as an RMarkdown document [@rmarkdown] at 
<https://github.com/cboettig/bad-forecast-good-decision>

@@

While in principle the utility could reflect many things, including the cost of fishing, market responses to supply and demand, the value of recreational fishing, the intrinsic value fish left in the sea [see @Halpern2013], for simplicity we will assume utility is merely a linear function of the harvest, i.e. a fixed price $p$ per kilogram of fish harvested: $U(X_t, H_t) = p H_t$. This problem is already well studied, and it is worth noting that even under such a pessimistic assumption, the optimal strategy still seeks to sustain the fish population indefinitely; as @Clark1973 shows for the deterministic function $f$ and @Reed1979 extended to the stochastic case. Given the function $f$ with known parameters, it is straight forward to determine the optimal harvest policy by stochastic dynamic programming (SDP, see @Mangel1985; @Marescot2013).

@@

```{r proper_scores, message = FALSE}
# Gneiting & Raferty (2007), eq24
scoring_fn <- function(x, mu, sigma){ -(mu - x )^2 / sigma^2  - log(sigma)}

@@

```{r plot_sims}
fig2a <- 
  sims %>% 
  group_by(model, time) %>%
  summarise(mean_state = mean(state), sd = sd(state), .groups = "drop") %>%
  filter(time < 25, model != "3") %>%
  ggplot(aes(time, mean_state)) + 
  geom_line(aes(col = model), lwd=1.5, show.legend = FALSE) +
    geom_ribbon(aes(ymin = mean_state - 2*sd, 
                    ymax = mean_state + 2*sd,
                    fill = model),
                alpha = 0.2, show.legend = FALSE) +
  ylab("state") + 
  ggtitle("A") + labs(subtitle = "Ecological performance")
```

@@

fig2b <- 
npv_df %>%
  filter(model != "3", time %in% seq(1,100, by = 5)) %>%
  ggplot(aes(time, mean_npv)) +
  geom_line(data = optimal, lwd = 1.5, col = "grey20") +
  geom_point(aes(col=model), size = 4, alpha = 0.8) + 
  ylab("Net present value") + xlab("time")  + 
  ggtitle("B.") + labs(subtitle = "Economic performance")

>>>>

---
title: 'Appendix for: "Bad Forecast, Good Decision"'
date: "`r Sys.Date()`"
author: Carl Boettiger
journal: Appendix
layout: 3p
bibliography: refs.bib
#output:
#  tufte::tufte_handout:
#    dev: cairo_pdf
#    latex_engine: xelatex
output: 
  hrbrthemes::ipsum_pdf:
    dev: cairo_pdf
    latex_engine: xelatex

@@

Here we provide annotated code necessary to completely reproduce all of the analysis presented in the main paper. This analysis is run in R [@R] uses `MDPtoolbox` [@MDPtoolbox] for solving Markov Decision Processes (MDP) using stochastic dynamic programming functionality, `expm` for matrix exponentials [@expm], and a few custom MDP functions provided by our package, `mdplearning` [@mdplearning].  We will also use `tidyverse`  packages for basic manipulation and plotting [@tidyverse]. This file is also available as an RMarkdown document [@rmarkdown] at  <https://github.com/cboettig/bad-forecast-good-decision>.

@@

While in principle the utility could reflect many things, including the cost of fishing, market responses to supply and demand, the value of recreational fishing, the intrinsic value fish left in the sea [see @Halpern2013], for simplicity we will assume utility is merely a linear function of the harvest quota set by the manager, i.e. a fixed price $p$ per kilogram of fish harvested: $U(X_t, H_t) = p \min(H_t, X_t)$ (noting that realized harvest cannot exceed the available stock). As units are considered already non-dimensionalized in this example, without loss of generality we will set $p=1$. This problem is already well studied, and it is worth noting that even under such a pessimistic assumption, the optimal strategy still seeks to sustain the fish population indefinitely; as @Clark1973 shows for the deterministic function $f$ and @Reed1979 extended to the stochastic case. Given the function $f$ with known parameters, it is straight forward to determine the optimal harvest policy by stochastic dynamic programming (SDP, see @Mangel1985; @Marescot2013).

@@

```{r proper_scores, message = FALSE}
# Gneiting & Raferty (2007), eq27
scoring_fn <- function(x, mu, sigma){ -(mu - x )^2 / sigma^2  - log(sigma)}

@@

```{r plot_sims}
fig2a <- 
  sims %>% 
  group_by(model, time) %>%
  summarise(mean_state = mean(state), sd = sd(state), .groups = "drop") %>%
  filter(time < 25, model != "3") %>%
  ggplot(aes(time, mean_state)) + 
  geom_line(aes(col = model), lwd=1.5, show.legend = FALSE) +
    geom_ribbon(aes(ymin = mean_state - 2*sd, 
                    ymax = mean_state + 2*sd,
                    fill = model),
                alpha = 0.2, show.legend = FALSE) +
  ylab("state") + 
  ggtitle("A") + labs(subtitle = "Ecological")
```

@@

fig2b <- 
npv_df %>%
  filter(model != "3", time %in% seq(1,100, by = 5)) %>%
  ggplot(aes(time, mean_npv)) +
  geom_line(data = optimal, lwd = 1.5, col = "grey20") +
  geom_point(aes(col=model), size = 4, alpha = 0.8) + 
  ylab("Net present value") + xlab("time")  + 
  ggtitle("B.") + labs(subtitle = "Economic")

### 2020-11-05 manuscript/manuscript.Rmd
here we go

A primary purpose of statistical analyses and ecological modeling is to make forecasts for the future [@Clark2001; @Gneiting2014].  Accurate forecasts are important both in assessing the accuracy of our models and understanding of natural processes, they also underpin policy and decision making [@Dietze2018].  Yet the model that leads to the best decisions is not always the model that makes the most accurate forecasts, as I illustrate here.  Ecological processes are intrinsically complex, so that even our best models can only ever be approximations of underlying processes.  In choosing the best model for decision-making, it can be more important to capture a single key feature of the process than it is to make the most accurate prediction about future states.  Even strictly proper scoring rules for probabilistic forecasts [@Gneiting2007] will not always select the best model to guide decision-making.  Surprisingly, this is true not only in conceptual models that may guide policy choices, but also when a decision policy is derived directly from a complex optimization routine of a probabilistic predictive model, such as the stochastic dynamic programming (SDP) algorithms used to derive optimal policies to sequential decision problems [@Marescot2013].  Here, I use a classical, well-understood example from fisheries management [@Schaefer1954; @Clark1973; @Reed1979] to illustrate both the paradox of how a model with the worst forecast provides the best decision outcomes, as well as show how we can avoid selecting models that are poorly suited for management by considering the management context more explicitly.

@@

For simplicity, I will focus on the classic case case of a single-species model whose population is observed annually without error in a stochastic but stationary environment without age structure [@Reed1979; @Clark1990; @Costello2016]. These are not necessary assumptions -- in fact, the more complex the models become, the easier it is find examples in which the best forecast does not produce the best decision.  Rather, using a simple model merely reflects the famous compromise of Richard Levins [@Levins1966] in choosing generality over precision.  More precisely, the decision problem in question can be stated as follows: The fish stock is observed to be in state `X_t` at time `t`, and is then subjected to some harvest `H_t` before recruiting new fish, subject to stochastic environmental noise `\xi_t`, to bring the stock to $X_t+1$,

@@

```{r figure1, fig.width=7, fig.height=5, fig.cap = "Forecast performance of each model.  Panels A, B: Step ahead predictions under unfished (A) and fished (B) scenarios. Error bars indicating the 95% confidence intervals around each prediction, while stars denote the observed value in that year.  Because the models make different decisions each year, the stock size in year 2, 3, etc under management of model 1 (blue stars) is different from that under model 2 (red stars).  Panels C, D: corresponding distribution of proper scores across all predictions. Higher scores are better."}
fig1ab + fig1cd + 
  plot_layout(widths =  c(3,1), guides = 'collect') + 
  theme(plot.margin = margin(0, 0, .1, .1, "cm"))
```

@@

```{r figure2, message = FALSE, fig.cap = "Ecological and economic performance of each forecast."}
fig2a + fig2b
```

@@

>>>>

A primary purpose of statistical analyses and ecological modeling is to make forecasts for the future [@Clark2001; @Gneiting2014].  Accurate forecasts are important both in assessing the accuracy of our models and understanding of natural processes, they also underpin policy and decision making [@Dietze2018].  Yet the model that leads to the best decisions is not always the model that makes the most accurate forecasts, as I illustrate here.  Ecological processes are intrinsically complex, so that even our best models can only ever be approximations of underlying processes.  Even strictly proper scoring rules for probabilistic forecasts [@Gneiting2007] will not always select the best model to guide decision-making.  Surprisingly, this can even happen when a decision is derived directly from a complex optimization routine of a probabilistic predictive models [@Marescot2013].  Here, I use a classic, well-understood example from fisheries management [@Schaefer1954; @Clark1973; @Reed1979] to illustrate both the paradox of how a model with the worst forecast provides the best decision outcomes, as well as show how we can avoid selecting models that are poorly suited for management by considering the management context more explicitly. These results underscore that in choosing the best model for decision-making, it can be more important to capture a single key feature of the process than it is to make the most accurate prediction about future states.

@@

For simplicity, I will focus on the classic case case of a single-species model whose population is observed annually without error in a stochastic but stationary environment without age structure [@Reed1979; @Clark1990; @Costello2016]. These are not necessary assumptions -- in fact, the more complex the models become, the easier it is find examples in which the best forecast does not produce the best decision.  Rather, using a simple model merely reflects the famous compromise of Richard Levins [@Levins1966] in choosing generality over precision.  More precisely, the decision problem in question can be stated as follows: The fish stock is observed to be in state $X_t$ at time $t$, and is then subjected to some harvest $H_t$ before recruiting new fish, subject to stochastic environmental noise $\xi_t$, to bring the stock to $X_t+1$,

@@

```{r figure1, fig.width=7, fig.height=5, fig.cap = "Forecast performance of each model.  Panels A, B: Step ahead predictions of stock size under unfished (A) and fished (B) scenarios. Error bars indicating the 95% confidence intervals around each prediction, while stars denote the observed value in that year.  Because the models make different decisions each year in the fished scenario, the observed stock size in year 2, 3, etc under the management of model 1 (blue stars) is different from that under model 2 (red stars).  Panels C, D: corresponding distribution of proper scores across all predictions (100 replicates of 100 timesteps). Higher scores are better, confirming that model 2 makes the better forecasts."}
fig1ab + fig1cd + 
  plot_layout(widths =  c(3,1), guides = 'collect') + 
  theme(plot.margin = margin(0, 0, .1, .1, "cm"))
```

@@

```{r figure2, message = FALSE, fig.cap = "Ecological and economic performance of each forecast. Harvest quotas derived from model 1 result in a significantly higher stock size than under model 2 (panel A).  Despite this less agressive harvest policy, economic returns under model 1 are substantially higher (panel B)."}
fig2a + fig2b
```

@@

# Methods

Stochastic transition matrices are defined for models 1-3 on a discrete grid of
240 possible states spaced uniformly from 0 to 24. A discrete action space 
enumerating possible harvest quotas is set to the same grid. The utility of a harvest quota $H_t$ given a population state $X_t$ is given by $U(X_t, H_t) = \min(X_t, H_t)$ (i.e. a fixed price for realized harvest). A modest discount of
$\gamma = 0.99$ allows comparisons to approaches that ignore [@Schaefer1954] or include [@Clark1973; @Reed1979] discounting; results are not sensitive to this choice. The optimal policy for each model is determined by stochastic dynamic programming [@Marescot2013].  Details of the implementation, including fully reproducible R code, have been included in the appendix.
